---
title: "Storm events in the US and the damages"
author: "Lihui LIU"
date: "26 juillet 2015"
output: html_document
---

SYNOPSIS: This report studies the human and economic loss caused by storm events throughout the US. The events are categorized following the norm defined by the US National Oceanic and Atmospheric Administration. The losses will be calculated for individual incidents of each type of event, as well as cumulatively for all incidents of each type of event. It will be shown that 

## Processing data


The raw data is provided by the US National Oceanic and Atmospheric Administration (NOAA). The following lines are for downloading and the data importing it into the data frame "df_Storm".

```{r, cache=TRUE}
# url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
# download.file(url,"storm_data.csv.bz2",method = "curl")
df_Storm <- read.csv("storm_data.csv.bz2")
```

The data needs cleaning up because the event names do not follow the the norm of NOAA. The latter classifies the hostile weather events in 48 types. The following lines of code creats a dataframe "index" storing these types 

```{r, cache=TRUE}
disasters <- c("Astronomical Low Tide","Avalanche","Blizzard","Costal Flood","Cold/Wind Chill","Debris Flow",
               "Dense Fog","Dense Smoke","Drought","Dust Devil","Dust Storm","Excessive Heat","Extreme Cold/Wind Chill",
               "Flash Flood","Flood","Freezing Fog","Frost/Freeze", "Funnel Cloud","Hail","Heat","Heavy Rain","Heavy Snow",
               "High Surf","High Wind","Hurricane/Typhoon","Ice Storm","Lakeshore Flood","Lake-Effect Snow",
               "Lightning","Marine Hail","Marine High Wind","Marine Strong Wind","Marine Thunderstorm Wind",
               "Rip Current","Seiche","Sleet","Storm Tide","Strong Wind","Thunderstorm Wind","Tornado",
               "Tropical Depression","Tropical Storm","Tsunami","Volcanic Ash","Waterspout","Wildfire",
               "Winter Storm","Winter Weather")
designators <- c("Z","Z","Z","Z","Z","C","Z","Z","Z","C","Z","Z","Z","C","C","Z","C","Z","C","Z","C","Z","Z","Z",
           "Z","Z","Z","Z","C","M","M","M","M","Z","Z","Z","Z","Z","C","C","Z","Z","Z","Z","M","Z","Z","Z")
index <- data.frame(ind=1:48, names = disasters, designators = designators)
```

The index data frame is
```{r,cache=TRUE}
print(index)
```

Then we can create a vector, whose length is the number of different events appearing in the raw data, and whose entries tell what type each event is, following NOAA's norm. The following code is for this purpose, where the procedure where some judgement of empirical nature is involved for some unclear descriptions. The vector created is "tidy_aux".

```{r,cache=TRUE}
EV_LVS <- levels(df_Storm$EVTYPE)
tidy_aux <- data.frame(lev= EV_LVS, lab = rep(0,length(EV_LVS)))

tidy_aux$lab[grep("summary",EV_LVS,ignore.case = TRUE)] <- -1

tidy_aux$lab[grep("low tide",EV_LVS,ignore.case = TRUE)] <- index[1,1] # Astronomical Low Tide 
tidy_aux$lab[c(18,19)] <- index[2,1] # Avalanche, excluding HEAVY SNOW/BLIZZARD/AVALANCHE

tidy_aux$lab[grep("FREEZING RAIN",EV_LVS,ignore.case = TRUE)] <- index[48,1] # Freezing rain as WINTER WEATHER

tidy_aux$lab[grep("blizzar",EV_LVS,ignore.case = TRUE)] <- index[3,1] # BLIZZARD

tidy_aux$lab[grep("coastal",EV_LVS,ignore.case = TRUE)] <- index[4,1] # COASTAL FLOOD
tidy_aux$lab[grep("erosi",EV_LVS,ignore.case = TRUE)] <- index[4,1] # COASTAL FLOOD
tidy_aux$lab[grep("blow-out",EV_LVS,ignore.case = TRUE)] <- index[4,1] # COASTAL FLOOD

tidy_aux$lab[grep("CSTL",EV_LVS,ignore.case = TRUE)] <- index[4,1]
tidy_aux$lab[grep("BEACH FLOOD",EV_LVS,ignore.case = TRUE)] <- index[4,1]

tidy_aux$lab[grep("wind chil",EV_LVS,ignore.case = TRUE)] <- index[5,1] 
# COLD/WIND CHILL (including also "EXTREME COLD WIND CHILL")
tidy_aux$lab[grep("low te",EV_LVS,ignore.case = TRUE)] <- index[5,1] 
tidy_aux$lab[c(80,146,147,414:416,880,881:883)] <- index[5,1] 
# COLD/WIND CHILL (including also "EXTREME COLD WIND CHILL")
tidy_aux$lab[c(65,66,72:77,129:140,148,
               535,536,548:550,569,603,896)] <- index[5,1] 

# No "DEBRIS FLOW" 6
tidy_aux$lab[grep("fog",EV_LVS,ignore.case = TRUE)] <- index[7,1] # DENSE FROG (temporarily containing freezing fog)
tidy_aux$lab[grep("smoke",EV_LVS,ignore.case = TRUE)] <- index[8,1] # DENSE SMOKE

tidy_aux$lab[grep("droug",EV_LVS,ignore.case = TRUE)] <- index[9,1] # DROUGHT, COVERING DROUGHT OF HEAT AND SNOW
tidy_aux$lab[grep("dry micro",EV_LVS,ignore.case = TRUE)] <- index[9,1] # DROUGHT, COVERING DROUGHT OF HEAT AND SNOW
tidy_aux$lab[c(11,98:100,109:112,137,397,401,495,497,554,555,884,888,922)] <- index[9,1] 
# DROUGHT, COVERING DROUGHT OF HEAT AND SNOW

tidy_aux$lab[c(25)] <- index[9,1] # BELOW NORMAL PRECIPITATION as DROUGHT

tidy_aux$lab[grep("dust de",EV_LVS,ignore.case = TRUE)] <- index[10,1] # DUST DEVIL
tidy_aux$lab[447] <- index[10,1] # DUST DEVIL

tidy_aux$lab[grep("dustst",EV_LVS,ignore.case = TRUE)] <- index[11,1] # DUST STORM
tidy_aux$lab[grep("dust st",EV_LVS,ignore.case = TRUE)] <- index[11,1] # DUST STORM
tidy_aux$lab[grep("blowing dust",EV_LVS,ignore.case = TRUE)] <- index[11,1] # DUST STORM
tidy_aux$lab[grep("SAHARAN",EV_LVS,ignore.case = TRUE)] <- index[11,1] # DUST STORM

tidy_aux$lab[c(130,142,282,580)] <- index[12,1] # EXCESSIVE HEAT, covering EXTREME HEAT and HEAT BURST, yielding to DROUGHT
tidy_aux$lab[c(14,26,27,45,46,47,78,79,86,87,120,
               141,143,144,145,191:196,273,482,963)] <- index[13,1] # EXTREME COLD/WIND CHILL

tidy_aux$lab[grep("FLASH FLOO",EV_LVS,ignore.case = TRUE)] <- index[14,1] # FLASH FLOOD
tidy_aux$lab[grep("FLOOD FL",EV_LVS,ignore.case = TRUE)] <- index[14,1] # FLASH FLOOD
tidy_aux$lab[grep("SNOWMELT FLOOD",EV_LVS,ignore.case = TRUE)] <- index[14,1] 
tidy_aux$lab[grep("STREET FLOOD",EV_LVS,ignore.case = TRUE)] <- index[14,1] 
tidy_aux$lab[grep("dam ",EV_LVS,ignore.case = TRUE)] <- index[14,1] 

tidy_aux$lab[grep("MINOR FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("MAJOR FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("LOCAL FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("RIVER FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("RURAL FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("RURAL FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("BREAKUP FLOODING",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[grep("TIDAL FLOOD",EV_LVS,ignore.case = TRUE)] <- index[15,1] # FLOOD
tidy_aux$lab[c(169,170,174,182,184,185,186,187,393,422,423,439,
               441:444,513,514,516:521,547,593)] <- index[15,1] # FLOOD

tidy_aux$lab[c(173,175,179,180,396,901,902,911,913,915,916,917)] <- index[14,1] 
#URBAN SMALL STREAM FLOOD considered as FLASH FLOOD

tidy_aux$lab[grep("FROST",EV_LVS,ignore.case = TRUE)] <- index[17,1] # FROST/FREEZE

tidy_aux$lab[c(67,68,214,215,216,217,219,220,448,595,931,932)] <- index[18,1] # FUNNEL CLOUD

tidy_aux$lab[grep("hail",EV_LVS,ignore.case = TRUE)] <- index[19,1] 
# HAIL, temporarily covering THUNDERSTORM and MARINE HAIL

tidy_aux$lab[c(100,275,277,278,280,400,413,556,557,558,885)] <- index[20,1] # HEAT, covering HEAT WAVE
tidy_aux$lab[grep("warm",EV_LVS,ignore.case = TRUE)] <- index[20,1]
tidy_aux$lab[grep("HIGH TEM",EV_LVS,ignore.case = TRUE)] <- index[20,1]

tidy_aux$lab[grep("high surf",EV_LVS,ignore.case = TRUE)] <- index[23,1] # HIGH SURF
tidy_aux$lab[grep("high wa",EV_LVS,ignore.case = TRUE)] <- index[23,1] # HIGH SURF
tidy_aux$lab[grep("swells",EV_LVS,ignore.case = TRUE)] <- index[23,1] # HIGH SURF
tidy_aux$lab[grep("surge",EV_LVS,ignore.case = TRUE)] <- index[23,1] # HIGH SURF
tidy_aux$lab[c(16,274,307,338:340,348,355,594,596,597)] <- index[23,1] # ROUGH SEA and HEAVY SURF as HIGH SURF

tidy_aux$lab[grep("high wind",EV_LVS,ignore.case = TRUE)] <- index[24,1] # HIGH WIND
tidy_aux$lab[c(8,107,108,181,225:228,233,236,237,241:243,347,494,526,
               640,669,952,953,959:962,965:968,970,985)]<- index[24,1] # GUSTY WIND, WIND DAMAGE, WHIRLWIND

tidy_aux$lab[grep("heavy rain",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN covering HIGH WIND and HIHG SURF
tidy_aux$lab[grep("URBAN FLOOD",EV_LVS,ignore.case = TRUE)] <- index[21,1] 
# URBAN FLOOD, considered as HEAVY RAIN, temporarily covering THUNDERSTORM
tidy_aux$lab[grep("STREAM FLOOD",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN
tidy_aux$lab[grep("STREAM  FLOOD",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN
tidy_aux$lab[grep("SML STREAM FLD",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN
tidy_aux$lab[grep("urban small",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN
tidy_aux$lab[grep("urban/small",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN
tidy_aux$lab[grep("urban and small",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN
tidy_aux$lab[grep("small stream and",EV_LVS,ignore.case = TRUE)] <- index[21,1] # HEAVY RAIN


tidy_aux$lab[c(132:134,225,239,240,285:287,308,309
               ,412,539:544,546,567,581,845,846)] <- index[21,1]

tidy_aux$lab[grep("sleet",EV_LVS,ignore.case = TRUE)] <- index[36,1] # SLEET
tidy_aux$lab[c(207,284,461,501:503,659)] <- index[36,1] # MIXED PRECIPITATION as SLEET

tidy_aux$lab[grep("heavy snow",EV_LVS,ignore.case = TRUE)] <- index[22,1] 
# HEAVY SNOW covering HIGH WIND temporarily covering SLEET
tidy_aux$lab[grep("snow squ",EV_LVS,ignore.case = TRUE)] <- index[22,1] 
tidy_aux$lab[c(13,42:44,135,150,283,344,418,430,431,537,628:633
               ,635,636,643,650:653,657,664,665,667,751,752)] <- index[22,1] 

tidy_aux$lab[grep("hurric",EV_LVS,ignore.case = TRUE)] <- index[25,1] # HURRICANE/TYPHOON
tidy_aux$lab[c(223,226,879)] <- index[25,1] # HURRICANE/TYPHOON
tidy_aux$lab[grep("ice st",EV_LVS,ignore.case = TRUE)] <- index[26,1] #ICE STORM covering FLASH FLOOD

tidy_aux$lab[c(439,441)] <- index[27,1] #LAKESHORE FLOOD

tidy_aux$lab[grep("effect snow",EV_LVS,ignore.case = TRUE)] <- index[28,1] #LAKE-EFFECT SNOW

tidy_aux$lab[c(4,463:465,468,469,470,471,474)] <- index[29,1] #LIGHTNING
tidy_aux$lab[grep("LIGNTNING",EV_LVS,ignore.case = TRUE)] <- index[28,1]

tidy_aux$lab[grep("marine hail",EV_LVS,ignore.case = TRUE)] <- index[30,1] # MARINE HAIL

tidy_aux$lab[grep("marine high",EV_LVS,ignore.case = TRUE)] <- index[31,1] # MARINE HIGH WIND
tidy_aux$lab[grep("marine strong",EV_LVS,ignore.case = TRUE)] <- index[32,1] # MARINE STRONG WIND

tidy_aux$lab[grep("tstm",EV_LVS,ignore.case = TRUE)] <- index[39,1] 
tidy_aux$lab[grep("thunderst",EV_LVS,ignore.case = TRUE)] <- index[39,1] 
tidy_aux$lab[grep("m wind",EV_LVS,ignore.case = TRUE)] <- index[39,1] 
# THUNDER STORM, temporarily covering MARINE THUNDERSTORM WIND
tidy_aux$lab[grep("microburst",EV_LVS,ignore.case = TRUE)] <- index[39,1]
tidy_aux$lab[grep("micoburst",EV_LVS,ignore.case = TRUE)] <- index[39,1]
tidy_aux$lab[grep("downburst",EV_LVS,ignore.case = TRUE)] <- index[39,1]

tidy_aux$lab[c(486,489,490)] <- index[33,1] # MARINE THUNDERSTORM WIND

tidy_aux$lab[grep("rip cur",EV_LVS,ignore.case = TRUE)] <- index[34,1]  # RIP CURRENT

tidy_aux$lab[grep("seich",EV_LVS,ignore.case = TRUE)] <- index[35,1]  # SEICH

tidy_aux$lab[c(432,675:680)] <- index[38,1]  # STRONG WIND

tidy_aux$lab[834:839] <- index[40,1]  # TORNADO
tidy_aux$lab[c(69,841,843,844)] <- index[40,1]  # TORNADO


tidy_aux$lab[grep("tropical dep",EV_LVS,ignore.case = TRUE)] <- index[41,1]  # TROPICAL DEPRESSION
tidy_aux$lab[grep("tropical st",EV_LVS,ignore.case = TRUE)] <- index[42,1]  # TROPICAL STORM

tidy_aux$lab[877] <- index[43,1]  # TSUNAMI

tidy_aux$lab[grep("volc",EV_LVS,ignore.case = TRUE)] <- index[44,1]

tidy_aux$lab[grep("water sp",EV_LVS,ignore.case = TRUE)] <- index[45,1]  # WATERSPOUT
tidy_aux$lab[936:945] <- index[45,1]  # WATERSPOUT
tidy_aux$lab[c(7,840,849)] <- index[45,1]  # WATERSPOUT

tidy_aux$lab[grep("wild",EV_LVS,ignore.case = TRUE)] <- index[46,1]  # WILD FIRE
tidy_aux$lab[grep("brush fi",EV_LVS,ignore.case = TRUE)] <- index[46,1] # brush fire counted as WILD FIRE
tidy_aux$lab[c(190,229,583)] <- index[46,1] # grass fire and forest fires counted as WILD FIRE

tidy_aux$lab[c(972,976)] <- index[47,1] # winter storm

tidy_aux$lab[grep("winter w",EV_LVS,ignore.case = TRUE)] <- index[48,1] # WINTER WEATHER
tidy_aux$lab[grep("wintry mix",EV_LVS,ignore.case = TRUE)] <- index[48,1] # WINTER WEATHER
tidy_aux$lab[grep("wintery mix",EV_LVS,ignore.case = TRUE)] <- index[48,1] # WINTER WEATHER
tidy_aux$lab[grep("winter mix",EV_LVS,ignore.case = TRUE)] <- index[48,1] # WINTER WEATHER
tidy_aux$lab[grep("light sn",EV_LVS,ignore.case = TRUE)] <- index[48,1] # WINTER WEATHER
tidy_aux$lab[grep("moderate sn",EV_LVS,ignore.case = TRUE)] <- index[48,1] # WINTER WEATHER
tidy_aux$lab[c(579)] <- index[48,1] # WINTER WEATHER
```


This done, we can add a new column to the raw data frame, indicating the type of event of each line conforming to NOAA norm. The new data frame is called "df_Storm2". Note that some unidentifiable events (such as WET) are purged.

```{r,cache=TRUE}
IND_VEC <- vector("numeric",dim(df_Storm)[1])

for(i in 1:length(IND_VEC)){
  indicator <- which(tidy_aux$lev == df_Storm$EVTYPE[i])
  IND_VEC[i] <- tidy_aux$lab[indicator]
}

IND_VEC <- IND_DF

IND_IND_VEC <- IND_VEC!=0 & IND_VEC!=-1

df_Storm2 <- cbind(df_Storm[IND_IND_VEC,],ST_EVTYPE = IND_VEC[IND_IND_VEC])
```

The final step of processing data is to create a data frame summarizing the damages of each of the 48 types of weather conditions. The data frame thus obtained is called "Final_DF".


```{r,cache=TRUE}
EV_Count <- Ave_Fatal <- Tot_Fatal <- Ave_Inj <- Tot_Inj <- Ave_PropDam <- Tot_PropDam <- Ave_CropDam <- Tot_CropDam <- vector("numeric",48)


for(i in 1:48) {
  temp_indicator <- df_Storm2$ST_EVTYPE==i
  temp_fat <- df_Storm2$FATALITIES[temp_indicator]
  temp_inj <- df_Storm2$INJURIES[temp_indicator]
  temp_PropDam <- df_Storm2$PROPDMG[temp_indicator]
  temp_CropDam <- df_Storm2$CROPDMG[temp_indicator]
  Ave_Fatal[i] <- mean(temp_fat)
  Tot_Fatal[i] <- sum(temp_fat)
  Ave_Inj[i] <- mean(temp_inj)
  Tot_Inj[i] <- sum(temp_inj)
  Ave_PropDam[i] <- mean(temp_PropDam)
  Tot_PropDam[i] <- sum(temp_PropDam)
  Ave_CropDam[i] <- mean(temp_CropDam)
  Tot_CropDam[i] <- sum(temp_CropDam)
  EV_Count[i] <- length(temp_fat)
}

Final_DF <- data.frame(EVInd = 1:48, EVTYPE = index[,3], Ave_Fatal = Ave_Fatal, Tot_Fatal = Tot_Fatal, 
                       Ave_Inj = Ave_Inj, Tot_Inj = Tot_Inj, 
                       Ave_PropDam = Ave_PropDam, Tot_PropDam = Tot_PropDam, 
                       Ave_CropDam = Ave_CropDam, Tot_CropDam = Tot_CropDam, Event_Count = EV_Count)

```

## Results 

```{r,cache=TRUE}

```

### abc
